---
title: "Explore data"
output: html_notebook
---


## Input data

This notebook has to be ran from the server as it directly access large dataset stored on disk.
Data are available on the `data2` disk. From the coral server, they can be accessed at `/export/isilon/datos2/satmo_analysis/`. The variable, collection and resolution is indicated in the file names. When a file has `.grd` + `.gri` extension (native format of the raster package), time information is directly written into the file. When the file is in `.tif`, it should come with a `.time` file, that is simply a text file containing the date for each layer.

## Origin of the data


Following download of L2 files from the oceancolor servers, data were processed using the following utilities:
    
- `timerange_L3m_process.py` for data masking and binning to a regular 2km grid
- `timerange_daily_composite.py` for combining data from the 3 sensors (aqua, terra, and viirs) into daily composites (mean value compositing)
- `timerange_time_composite.py` to create timcomposites (8 days and 16 days) from the daily composites
- A custom stacking script using `rasterio` to stack layers in chronological order (see `python` directory)


## Global variables

```{r, echo=FALSE, warning=FALSE}
library(raster)
library(rasterVis)
library(rgdal)
library(magrittr)
library(ggplot2)
library(zoo)
path <- '/export/isilon/datos2/satmo_analysis'
```



## Visualize chlorophyll a

### Load data and assign time to multitemporal raster object

```{r}
chl_a_brick <- file.path(path, 'stack.L3m_8DAY_CHL_chlor_a_2km.tif') %>%
    brick()
chl_a_time <- file.path(path, 'stack.L3m_8DAY_CHL_chlor_a_2km.time') %>%
    scan(what = character()) %>%
    as.Date()
chl_a_brick <- setZ(chl_a_brick, chl_a_time)
```


### Visualize a layer


```{r}
layer_nb <- 400
title <- sprintf('Chlor_a, 8 day composite, %s', getZ(chl_a_brick)[layer_nb])
levelplot(chl_a_brick, layers=layer_nb, zscaleLog=TRUE, maxpixels=1000000, margin = FALSE, main=title, col.regions = colorRampPalette(c("#00007F", "blue", "#007FFF", "cyan", "#7FFF7F", "yellow", "#FF7F00", "red", "#7F0000"))(250))
```


### Extract and visualize temporal profiles of individual pixels

```{r}
source('R/tsExtract.R')
ts <- tsExtractCoords(chl_a_brick, long=-88, lat=22)
ggplot(fortify(ts)) +
    geom_point(aes(Index, ts)) +
    xlab('Time') +
    ylab('Chlor_a')
```


We can add a bit of temporal context by plotting a fitted harmonic (or harmonic trend) model together with the observations.

```{r}
source('R/harmonFit.R')
df <- harmonFit(ts) # See the sourced function for additional parameters (order, formula, fit)
ggplot(df) +
    geom_point(aes(dates, response)) +
    geom_line(aes(dates, prediction),  color = 'green') +
    xlab('Time') +
    ylab('Chlor_a')
```

### Save data

We can save the data as an example dataset for further analysis.

```{r}
saveRDS(df, file = '../data/chlor_a_ts.rds')
```













