---
title: "Outlier detection"
output: html_notebook
---

This notebook prototypes a simple system for detecting outliers in a time-series. Such approach could later be used (in combination with others) for the detecion of events like algae blooms.

## Packages and data

Refer to notebook `01-explore-satmo-data` to know where the time-series loaded below come from. It contains observations of chlorophyll_a, predicted values from a hrmonic trend model fit and residuals.

```{r}
library(ggplot2)
df <- readRDS('../data/chlor_a_ts.rds')
source('R/outlierDetector.R')
```

The `outlierDetector` computes the interquartile Range (IQR) on the residuals of the a fitted model (harmonic trend in the test dataset). A sensitivity value is applied to this interquantile range to determine boundaries that define whether a residual value is an outlier or not. By default the `outlierDetector` uses a sensitivity value of 1.5. This means that if an observation has an outlier value that exceeds `1.5 * first or third quartile * IRQ`, it is considered to be an outlier.

```{r}
df2 <- outlierDetector(df)
ggplot(df2) +
    geom_point(aes(dates, response, color = isOutlier)) +
    guides(color = "none") +
    geom_line(aes(dates, prediction),  color = 'green') +
    xlab('Time') +
    ylab('Chlor_a')
```

In the above figure, blue dots correspond to outliers, relative to the harmonic trend fit. Sensitivity value can be adjusted to increase the non-outlier range.

```{r}
df3 <- outlierDetector(df, sensitivity = 3)
ggplot(df3) +
    geom_point(aes(dates, response, color = isOutlier)) +
    guides(color = "none") +
    geom_line(aes(dates, prediction),  color = 'green') +
    xlab('Time') +
    ylab('Chlor_a')
```

Increasing the sensitivity makes the 'valid range' wider, and results in the detection of less outliers. It's a bit counter intuitive, but it's like that!

